Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-12 talk/pepm2011/paper.tex
flesh out the benchmarking and related work sections a bit more. add some acknowlegdements

\keywords{XXX}%

@@

Benchmarks from the Computer Language Benchmark Game are: fannkuch, nbody,
meteor-contest, spectral-norm.

\textbf{crypto\_pyaes}: AES implementation.
\textbf{django}: The templating engine of the Django web
framework\footnote{\texttt{http://www.djangoproject.com/}}.
\textbf{go}: A Monte-Carlo Go
AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/disco-elegant-python-go-player.html}}.
\textbf{html5lib}: HTML5 parser
\textbf{pyflate-fast}: BZ2 decoder
\textbf{raytrace-simple}: ray tracer
\textbf{richards}: The Richards benchmark \cite{XXX}
\textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
\textbf{telco}: A Python version of the Telco decimal
benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a pure
Python decimal floating point implementation.
\textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
framework\footnote{\texttt{http://twistedmatrix.com/trac/}}.

@@

@@

>>>>

\keywords{Tracing JIT, Partial Evaluation, Optimization}

@@

To evaluate the effectiveness of our allocation removal algorithm, we look at
the effectiveness when used in the tracing JIT of PyPy's Python interpreter. The
benchmarks we used are small-to-medium Python programs, some synthetic
benchmarks, some real applications.

Some of them are from the Computer Language Benchmark
Game\footnote{\texttt{http://shootout.alioth.debian.org/}}: \textbf{fannkuch},
\textbf{nbody}, \textbf{meteor-contest}, \textbf{spectral-norm}.

Furthermore there are the following benchmarks:
\begin{itemize}
    \item \textbf{crypto\_pyaes}: AES implementation.
    \item \textbf{django}: The templating engine of the Django web
    framework\footnote{\texttt{http://www.djangoproject.com/}}.
    \item \textbf{go}: A Monte-Carlo Go
    AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
    \item \textbf{html5lib}: HTML5 parser
    \item \textbf{pyflate-fast}: BZ2 decoder
    \item \textbf{raytrace-simple}: ray tracer
    \item \textbf{richards}: The Richards benchmark \cite{XXX}
    \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
    \item \textbf{telco}: A Python version of the Telco decimal
    benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
    using a pure Python decimal floating point implementation.
    \item \textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
    framework\footnote{\texttt{http://twistedmatrix.com/}}.
\end{itemize}

We evaluate the allocation removal algorithm along two lines: First we want to
know how many allocations could be optimized away. On the other hand, we want
to know how much the run times of the benchmarks is improved.

For the former we counted the occurring operations in all generated traces
before and after the optimization phase for all benchmarks. The results can be
seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
little as XXX percent of allocation operations in the benchmarks. All benchmarks
taken together, the optimization removes XXX percent of allocation operations.

\begin{figure*}
\begin{tabular}{lrrrrrrrrrrrrrrrrrrrrrrr}
	&\#loops	&new	&	&	&get	&	&	&set	&	&	&guard	&	&	&rest	&	&	&	&	&	&total	&	&\\
crypto\_pyaes	&75	&1663	&446	&27\%	&18842	&2127	&11\%	&14552	&2020	&14\%	&8952	&235	&3\%	&53339	&7794	&15\%	&44009	&4828	&11\%	&97348	&12622	&13\%\\
django	&53	&453	&155	&34\%	&11970	&1457	&12\%	&8738	&991	&11\%	&4093	&275	&7\%	&33988	&4076	&12\%	&25254	&2878	&11\%	&59242	&6954	&12\%\\
fannkuch	&40	&161	&76	&47\%	&574	&306	&53\%	&278	&139	&50\%	&1121	&202	&18\%	&2581	&1685	&65\%	&2134	&723	&34\%	&4715	&2408	&51\%\\
go	&510	&5604	&1651	&29\%	&112403	&12612	&11\%	&88439	&11915	&13\%	&51621	&3194	&6\%	&300831	&30834	&10\%	&258067	&29372	&11\%	&558898	&60206	&11\%\\
html5lib	&416	&5939	&1793	&30\%	&241402	&15863	&7\%	&178723	&13960	&8\%	&58486	&3037	&5\%	&682654	&40341	&6\%	&484550	&34653	&7\%	&1167204	&74994	&6\%\\
meteor-contest	&56	&221	&141	&64\%	&2326	&606	&26\%	&2047	&645	&32\%	&1040	&169	&16\%	&7018	&2250	&32\%	&5634	&1561	&28\%	&12652	&3811	&30\%\\
nbody	&10	&84	&48	&57\%	&287	&159	&55\%	&121	&74	&61\%	&411	&90	&22\%	&986	&737	&75\%	&903	&371	&41\%	&1889	&1108	&59\%\\
pyflate-fast	&161	&1878	&723	&38\%	&22815	&2592	&11\%	&18415	&3690	&20\%	&8291	&628	&8\%	&64649	&11327	&18\%	&51399	&7633	&15\%	&116048	&18960	&16\%\\
raytrace-simple	&117	&1906	&359	&19\%	&52995	&2202	&4\%	&39518	&1823	&5\%	&13451	&447	&3\%	&140723	&6533	&5\%	&107870	&4831	&4\%	&248593	&11364	&5\%\\
richards	&79	&375	&191	&51\%	&27507	&2095	&8\%	&21217	&2004	&9\%	&3972	&332	&8\%	&76751	&6171	&8\%	&53071	&4622	&9\%	&129822	&10793	&8\%\\
spambayes	&267	&2413	&505	&21\%	&53327	&3033	&6\%	&40938	&2428	&6\%	&19343	&718	&4\%	&143575	&10322	&7\%	&116021	&6684	&6\%	&259596	&17006	&7\%\\
spectral-norm	&35	&266	&88	&33\%	&3004	&398	&13\%	&2544	&414	&16\%	&1084	&78	&7\%	&4803	&1184	&25\%	&6898	&978	&14\%	&11701	&2162	&18\%\\
telco	&42	&660	&70	&11\%	&20716	&424	&2\%	&14444	&608	&4\%	&6574	&56	&1\%	&53674	&1224	&2\%	&42394	&1158	&3\%	&96068	&2382	&2\%\\
twisted-names	&201	&2706	&317	&12\%	&64438	&2795	&4\%	&47960	&1968	&4\%	&24091	&555	&2\%	&171582	&6828	&4\%	&139195	&5635	&4\%	&310777	&12463	&4\%\\
total	&2062	&24329	&6563	&27\%	&632606	&46669	&7\%	&477934	&42679	&9\%	&202530	&10016	&5\%	&1737154	&131306	&8\%	&1337399	&105927	&8\%	&3074553	&237233	&8\%\\
\end{tabular}
\caption{Number of Operations Before and After Optimization}
\label{fig:numops}
\end{figure*}

In addition to the count of operations we also performed time measurements. The
machine the benchmarks were run on is XXX. We compared the performance of
various Python implementations on the benchmarks. As a baseline, we used the
standard Python implementation in C, called
CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. Furthermore we compared against Psyco \cite{XXX}, an extension to
CPython which is a just-in-time compiler that produces machine code at run-time.
It is not based on traces. Of PyPy's Python interpreter we used three versions,
one without a JIT, one including the JIT but not using the allocation removal
optimization, and one using the allocation removal optimizations.

All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{XXX}. The results are reported in Figure~\ref{fig:times}.
With the optimization turned on, PyPy's Python interpreter outperforms CPython
in all benchmarks except spambayes (which heavily relies on regular expression
performance). All benchmarks are improved by the allocation removal
optimization, some as much as XXX. XXX Psyco

XXX runtimes of the algorithm somehow?

\begin{figure*}
\begin{tabular}{lrrrr}
		&CPython	&JIT no optimizations	&JIT allocation removal	&JIT full\\
crypto\_pyaes	&1958.40 $\pm$ 0.32	&782.14 $\pm$ 3.08	&150.86 $\pm$ 3.62	&125.68 $\pm$ 3.02\\
django		&630.91 $\pm$ 0.13	&365.90 $\pm$ 1.57	&124.41 $\pm$ 0.88	&117.36 $\pm$ 0.91\\
fannkuch	&1344.15 $\pm$ 0.36	&378.18 $\pm$ 0.36	&309.04 $\pm$ 1.56	&299.12 $\pm$ 0.25\\
go		&595.31 $\pm$ 0.51	&1067.26 $\pm$ 12.20	&121.76 $\pm$ 4.15	&125.57 $\pm$ 3.90\\
html5lib	&8678.67 $\pm$ 23.37	&17665.16 $\pm$ 2695.45	&6739.95 $\pm$ 1225.25	&6699.24 $\pm$ 1295.92\\
meteor-contest	&241.53 $\pm$ 0.09	&291.11 $\pm$ 0.86	&274.09 $\pm$ 0.37	&272.77 $\pm$ 0.30\\
nbody		&396.55 $\pm$ 0.24	&84.77 $\pm$ 0.14	&69.01 $\pm$ 0.08	&68.71 $\pm$ 0.09\\
pyflate-fast	&1991.33 $\pm$ 1.90	&1818.70 $\pm$ 4.30	&1062.03 $\pm$ 4.56	&1052.02 $\pm$ 3.24\\
raytrace-simple	&1598.84 $\pm$ 1.07	&1126.93 $\pm$ 11.42	&93.45 $\pm$ 4.75	&92.17 $\pm$ 4.15\\
richards	&216.47 $\pm$ 0.20	&169.84 $\pm$ 1.21	&11.81 $\pm$ 0.09	&11.90 $\pm$ 0.29\\
spambayes	&193.30 $\pm$ 0.03	&331.37 $\pm$ 14.93	&238.83 $\pm$ 6.10	&237.23 $\pm$ 5.97\\
spectral-norm	&375.91 $\pm$ 0.08	&221.24 $\pm$ 0.07	&31.96 $\pm$ 0.06	&28.79 $\pm$ 0.09\\
telco		&789.67 $\pm$ 1.14	&622.67 $\pm$ 1.86	&155.33 $\pm$ 1.82	&154.67 $\pm$ 1.82\\
twisted\_names	&6.61 $\pm$ 0.00	&8.85 $\pm$ 0.09	&4.35 $\pm$ 0.05 &4.35 $\pm$ 0.04\\
\end{tabular}
\caption{Benchmark Times}
\label{fig:times}
\end{figure*}

@@

Escape analysis: XXX

partial evaluation:

Prolog

partially static data structures: kenichi asai's thesis?

hint in first js paper by michael franz \cite{mason_chang_efficient_2007}

SPUR, a tracing JIT for C# seems to be able to remove allocations in a similar
way to the approach described here, as hinted at in the technical report
\cite{XXX}. However, no details for the approach and its implementation are
given.

@@

\section*{Acknowledgements}

The authors would like to thank Stefan Hallerstede and Thomas Stiehl for
fruitful discussions during the writing of the paper.
